\textbf{The Structure of Chaos: An Empirical Comparison of Fractal Physiology Complexity Indices using NeuroKit2}

### 2022-07-20 analysis.Rmd

```{r time, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.cap="Median computation time difference between the different complexity indices algorithms, as well as variability as a function of signal lengths (represented by different line colors). The indices are grouped in sections (background color) according to their median computation time. Note that the time is expressed in arbitrary units as it is intended to convey differences, as the actual time would depend on the system specifications.", fig.height=10*1.25, fig.width=16*1.25, cache=cache}
dfsummary <- df |>
  filter(!Index %in% c("SD", "Length", "Noise", "Random", "Frequency")) |>
  mutate(Duration = Duration * 10000) |>
  group_by(Index, Length) |>
  summarize(
    CI_low = median(Duration) - mad(Duration),
    CI_high = median(Duration) + mad(Duration),
    Duration = median(Duration)
  )
dfsummary$CI_low[dfsummary$CI_low < 0] <- 0

@@

The second factor was strongly loaded by by indices that measure the feature-richeness of the signal's system (as most of them operate on a state-space decomposition). It was found to be positively related to *SVDEn* and the Kozachenko-Leonenko differential entropy (*KLEn*), and negatively to the RQA *Recurrence Rate* and *Hjorth* Complexity.

The third factor was loaded predominantly by permutation-based metrics (*PEn*, *WPEn*, *BubblEn*, etc.). The fourth factor included multiscale indices, such as *MSWPEn*. The fifth factor was strongly loaded by signal *length*, and thus might not capture features of complexity *per se*. Indices with the most relation to it were indices generally known to be sensitive to signal length, such as *ApEn*. The sixth factor was loaded by indices in which the signal or the Poincaré plot was discretized via binning or gridding, respectively. The seventh factor was loaded by sign-based entropy increments, and the eights by multiscale *IncrEn* and multiscale *PLZC*. The ninth factor was loaded by *EnofEn* and Kolmogorov Entropy (*K2En*). The tenth factor was loaded positively by the amount of noise, and negatively by multifractal indices such as *MFDFA (Width)*, suggesting a sensitivity to regularity. Finally, as a manipulation check for our factorization method, the random vector did not load unto any factors.

@@

For illustration purposes, we represented the correlation matrix as a connectivity graph (see **Figure 5**). We then ran a hierarchical clustering (with a Ward D2 distance) to provide additional information about the groups discussed above. Indeed, while the factor analysis will predominantly show indices that are the most representative of a given latent dimension, clustering will construct groups based on the multidimensional profile (what dimensions a given index loads positively on, and what other does it load negatively on). This allowed us to refined our recommendations for complimentary complexity indices (see **Figure 6**).

@@

The selection of a subset of indices was based on a set of considerations: 1) high loadings on one predominant latent dimension, with additional attention to the pattern of secondary loadings. For instance, an index with a positive factor 1 loading and a negative factor 2 loading could complement another index with a similar factor 1 loading, but a positive factor 2 loading. This was facilitated by 2) the hierarchical clustering dendrogram (see **Figure 6**), with which we attempted to extract indices from each (meaningful) higher order clusters. Items related to clusters that we determine as being largely explained by noise, length or other artifacts were omitted. 3) A preference for indices with relatively shorter computation times. This yielded a selection of 12 indices. Next, we computed the cumulative variance explained of this selection in respect to the entirety of indices examined, and derived the optimal order to maximize the variance explained (see **Figure 7**). The 12 included indices, representing `r insight::format_value(max(optim$VarExplained), as_percent=TRUE)` of the variance of the whole dataset, were:

@@

As the span and application of complexity science grows, a systematic approach to compare their "performance" becomes necessary to reinforce the clarity and structure of the field. The term *performance* used here is to be understood in a relative sense, as any such endeavor faces the "hard problem" of complexity science: various objective properties of signals [e.g., short-term vs. long-term variability, auto-correlation, information, randomness, @namdari2019; @xiong2017] participate in forming overarching concepts such as "complex" and "chaotic". Indices that are sensitive to some of these objective properties are thus conceptually linked through such over-arching frameworks. However, it remains unclear how these high-level concepts transfer back, in a top-down fashion, into a combination of lower-level features. As such, it is conceptually complicated to benchmark complexity measures against "objectively" complex vs. non-complex signals. In other words, while we are aware that different objective signal characteristics can contribute to the "complexity" of a signal, there is not a one-to-one correspondence between the latter and the former.

@@

Our results empirically confirm the plurality of underlying components of complexity (although it is here defined somewhat circularly as what is measured by complexity indices), and more importantly show that complexity indices vary in their sensitivity to various orthogonal latent dimensions. However, the mostly descriptive interpretation of most of these dimensions is a limitation of the present investigation, and future studies are needed to investigate and discuss them in greater depth (for instance, by modulating specific properties of signals and measuring their impact on these latent dimensions).

>>>>

```{r time, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.cap="Median computation time difference between the different complexity indices algorithms, as well as variability as a function of signal lengths (represented by different line colors). The indices are grouped in sections (background color) according to their median computation time. Note that the time is expressed in arbitrary units as it is intended to convey differences, since the actual time would depend on the system specifications.", fig.height=10*1.25, fig.width=16*1.25, cache=cache}
dfsummary <- df |>
  filter(!Index %in% c("SD", "Length", "Noise", "Random", "Frequency")) |>
  mutate(Duration = Duration * 10000) |>
  group_by(Index, Length) |>
  summarize(
    CI_low = median(Duration) - mad(Duration),
    CI_high = median(Duration) + mad(Duration),
    Duration = median(Duration)
  )
dfsummary$CI_low[dfsummary$CI_low < 0] <- 0

@@

The second factor was strongly loaded by indices that measure the feature-richness of the signal's system (as most of them operate on a state-space decomposition). It was found to be positively related to *SVDEn* and the Kozachenko-Leonenko differential entropy (*KLEn*), and negatively to the RQA *Recurrence Rate* and *Hjorth* Complexity.

The third factor was loaded predominantly by permutation-based metrics (*PEn*, *WPEn*, *BubblEn*, etc.). The fourth factor included multiscale indices, such as *MSWPEn*. The fifth factor was strongly loaded by signal *length*, and thus might not capture features of complexity *per se*. Indices with the most relation to it were indices generally known to be sensitive to signal length, such as *ApEn*. The sixth factor was loaded by indices in which the signal or the Poincaré plot was discretized via binning or gridding, respectively. The seventh factor was loaded by sign-based entropy increments, and the eighth by multiscale *IncrEn* and multiscale *PLZC*. The ninth factor was loaded by *EnofEn* and Kolmogorov Entropy (*K2En*). The tenth factor was loaded positively by the amount of noise, and negatively by multifractal indices such as *MFDFA (Width)*, suggesting a sensitivity to regularity. Finally, as a manipulation check for our factorization method, the random vector did not load unto any factors.

@@

For illustration purposes, we represented the correlation matrix as a connectivity graph (see **Figure 5**). We then ran a hierarchical clustering (with a Ward D2 distance) to provide additional information about the groups discussed above. Indeed, while the factor analysis will predominantly show indices that are the most representative of a given latent dimension, clustering will construct groups based on the multidimensional profile (what dimensions a given index loads positively on, and what other does it load negatively on). This allowed us to refine our recommendations for complimentary complexity indices (see **Figure 6**).

@@

The selection of a subset of indices was based on a set of considerations: 1) high loadings on one predominant latent dimension, with additional attention to the pattern of secondary loadings. For instance, an index with a positive factor 1 loading and a negative factor 2 loading could complement another index with a similar factor 1 loading, but a positive factor 2 loading. This was facilitated by 2) the hierarchical clustering dendrogram (see **Figure 6**), with which we attempted to extract indices from each (meaningful) higher order clusters. Items related to clusters that we determine as being largely explained by noise, length or other artifacts were omitted. 3) A preference for indices with relatively shorter computation times. This yielded a selection of 12 indices. Next, we computed the cumulative variance explained by this selection in respect to the entirety of indices examined, and derived the optimal order to maximize the variance explained (see **Figure 7**). The 12 included indices, representing `r insight::format_value(max(optim$VarExplained), as_percent=TRUE)` of the variance of the whole dataset, were:

@@

As the span and application of complexity science grows, a systematic approach to compare their "performance" becomes necessary to reinforce the clarity and structure of the field. The term *performance* used here is to be understood in a relative sense, as any such endeavor faces the "hard problem" of complexity science: various objective properties of signals [e.g., short-term vs. long-term variability, auto-correlation, information, randomness, @namdari2019; @xiong2017] participate in forming overarching concepts such as "complex" and "chaotic". Indices that are sensitive to some of these objective properties are thus conceptually linked through such overarching frameworks. However, it remains unclear how these high-level concepts transfer back, in a top-down fashion, into a combination of lower-level features. As such, it is conceptually complicated to benchmark complexity measures against "objectively" complex vs. non-complex signals. In other words, while we are aware that different objective signal characteristics can contribute to the "complexity" of a signal, there is not a one-to-one correspondence between the latter and the former.

@@

Our results empirically confirm the plurality of underlying components of complexity (although it is here defined somewhat circularly as what is measured by complexity indices), and more importantly show that complexity indices vary in their sensitivity to various orthogonal latent dimensions. However, the mostly descriptive interpretation of these dimensions is a limitation of the present investigation, and future studies are needed to investigate and discuss them in greater depth (for instance, by modulating specific properties of signals and measuring their impact on these latent dimensions).
